---
title: "# KNN and Tree Based Ensemble Models"
author: "Chien-Lan Hsueh"
date: "2022-07-04"
output:
  rmdformats::robobook: 
    theme: cerulean
    highlight: haddock
    code_folding: none
    df_print: paged
  github_document:
    toc: true
    df_print: kable
    html_preview: false
    math_method: webtex
  pdf_document:
    latex_engine: xelatex
    highlight: haddock
    df_print: tibble
  html_document:
    toc: true
    theme: cerulean
    highlight: haddock
    code_folding: none
    df_print: paged
---

## Project Goal

In this report, we conduct an exploratory data analysis (EDA) on the [Heart Failure Prediction Dataset](https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction). Then we fit KNN and ensemble models with repeated cross-validation on training data Test the models by finding the confusion matrix on the test data:

- K-Nearest Neighbors (KNN)
- Ensemble Models
  - Classification Tree
  - Bagged Tree
  - Random Forests
  - Generalized Boosted Regression Models

## Set up: Packages and Helper Functions

In this task, we will use the following packages:

- `here`: enables easy file referencing and builds file paths in a OS-independent way
- `stats`: loads this before loading `tidyverse` to avoid masking some `tidyverse` functions
- `tidyverse`: includes collections of useful packages like `dplyr` (data manipulation), `tidyr` (tidying data),  `ggplots` (creating graphs), etc.
- `glue`: embeds and evaluates R expressions into strings to be printed as messages
- `scales`: formats and labels scales nicely for better visualization
- `caret`: training and plotting classification and regression models
- `rpart`: recursive partitioning for classification, regression and survival trees. 
- `randomForest`: classification and regression based on a forest of trees using random inputs.
- `gbm`: generalized boosted regression models

```{r, include=FALSE}
if (!require("pacman")) utils::install.packages("pacman", dependencies = TRUE)

pacman::p_load(
  here,
  stats,
  tidyverse,
  glue,
  skimr,
  caret,
  rpart, randomForest, gbm
)
```

First, we define a helper functions to reduce repeating codes. This function trains models with cross validation to tune parameters and apply the best model on test set to see its performance.

> Arguments:
>
> - `form`: formula
> - `df_train`: training set
> - `df_test`: test set
> - `method`: classification or regression model to use
> - `trControl`: a list of values that define how train acts
> - `tuneGrid`: a data frame with possible tuning values
> - `plot`: whether to plot parameter and metric
> - `...`: arguments passed to the classification or regression routine
>
> Returned Value: a confusion matrix

```{r}
fit_model <- function(form, df_train, df_test, method, trControl, tuneGrid = NULL, plot = T, ...) {
  # train model
  fit <- train(
    form = form,
    data = df_train,
    method = method,
    preProcess = c("center", "scale"),
    trControl = trControl,
    tuneGrid = tuneGrid, ...)
  
  # print the best tune if there is a tuning parameter
  if(is.null(tuneGrid)){
    print("No tuning parameter")
  } else {
    # print the best tune 
    print("The best tune is found with:")
    print(glue("\t{names(fit$bestTune)} = {fit$bestTune[1,]}"))
  
    if(plot){
      # get model info
      model <- fit$modelInfo$label
      parameter <- fit$modelInfo$parameters$parameter
      description <- fit$modelInfo$parameters$label
      
      # plot parameter vs metrics
      p <- fit$results %>% 
        rename_at(1, ~"x") %>% 
        pivot_longer(cols = -1, names_to = "Metric") %>% 
        ggplot(aes(x, value, color = Metric)) +
        geom_point() +
        geom_line() +
        facet_grid(rows = vars(Metric), scales = "free_y") +
        labs(
          title = glue("{model}: Hyperparameter Tuning"),
          x = glue("{parameter} ({description})")
        )
      print(p)
    }
  }
  
  # make prediction on test set
  pred <- predict(fit, newdata = df_test)

  # confusion matrix
  cfm <- confusionMatrix(df_test[,1] %>% as_vector(), pred)
  
  # print confusion matrix and accuracy
  print("Confusion table:")
  print(cfm$table)
  print(glue("Accuracy = {cfm$overall['Accuracy']}"))

  # return the confusion matrix
  return(cfm)
}
```

## Data

The [Heart Failure Prediction Dataset](https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction) gives information about whether or not someone has heart disease along with different measurements about that person’s health. A local copy is saved in the `data` folder. Since the original column names contain space and special characters, and are long as well, we rename the columns when we read the data in.

```{r}
df_raw <- read_csv(here("data", "heart.csv"))
 
# show the raw data
df_raw

# check structure
str(df_raw)

# check if any missing values
anyNA(df_raw)
```

Next, we will prepare the data:

- remove `ST_Slope`
- convert the three categorical predictors `Sex`, `ChestPainType` and `RestingECG` to factors
- convert the response `HeartDisease` to a factor and relocate it to the first column

```{r}
df <- df_raw %>% 
  select(-ST_Slope) %>% 
  mutate(
    Sex = factor(Sex),
    ChestPainType = factor(ChestPainType),
    RestingECG = factor(RestingECG ),
    ExerciseAngina = factor(ExerciseAngina),
    HeartDisease = if_else(HeartDisease == 1, "Heart Disease", "Normal") %>% factor()
  ) %>% 
  relocate(HeartDisease)

# show the data frame
df

# quick summaries of numeric and categorical variables
skim(df)
```

## Split the Data

We will use `caret::createDataPartition()` to create a 80/20 split of training and test sets.

```{r}
set.seed(2022)

# split data
trainIndex <- createDataPartition(df$HeartDisease, p = 0.8, list = FALSE)
df_train <- df[trainIndex, ]
df_test <- df[-trainIndex, ]
```

## Part 1: kNN

To use kNN, we will need to encode the categorical predictors.

```{r, warning=FALSE}
# one-hot encode
dummies_model <- dummyVars(HeartDisease ~ ., data = df)
df_encoded <- bind_cols(df[, 1], predict(dummies_model, newdata = df))

# show the encoded data frame
df_encoded

# do the same split pattern for training and test sets
df_encoded_train <- df_encoded[trainIndex, ]
df_encoded_test <- df_encoded[-trainIndex, ]
```

Train a kNN model on the standardize data (centered and scaled) using repeated cross validation (10 folds, 3 repeats) to determine the best parameter $k = {1, 2,... , 40}$. 

```{r}
# train a kNN model with cv and apply the best model on test set
cfm_knn <- fit_model(
  HeartDisease ~ ., df_encoded_train, df_encoded_test, "knn", 
  trControl = trainControl(method = "repeatedcv", number = 10, repeats = 3),
  tuneGrid = expand.grid(k = 1:40))
```

## Part 2: Ensemble

Train a classification tree model on the standardize data (centered and scaled) using repeated cross validation (5 folds, 3 repeats) to determine the best parameter $\text{cp} = {0, 0.001, 0.002,... ,0.100}$.

```{r}
# train a classification tree model with cv and apply the best model on test set
cfm_tree <- fit_model(
  HeartDisease ~ ., df_train, df_test, "rpart", 
  trControl = trainControl(method = "repeatedcv", number = 5, repeats = 3),
  tuneGrid = expand.grid(cp = (1:100)/1000))
```

Train a bagged tree model on the standardize data (centered and scaled) using repeated cross validation (5 folds, 3 repeats).

```{r}
# train a bagged tree model with cv and apply the best model on test set
cfm_bag <- fit_model(
  HeartDisease ~ ., df_train, df_test, "treebag", 
  trControl = trainControl(method = "repeatedcv", number = 5, repeats = 3), plot = FALSE)
```

Train a random forest model on the standardize data (centered and scaled) using repeated cross validation (5 folds, 3 repeats) to determine the best parameter $\text{mtry} = {1, 2, ..., 15}$.

```{r}
# train a random forest model with cv and apply the best model on test set
cfm_rf <- fit_model(
  HeartDisease ~ ., df_train, df_test, "rf", 
  trControl = trainControl(method = "repeatedcv", number = 5, repeats = 3),
  tuneGrid = expand.grid(mtry = 1:15))
```

Train a boosted tree model on the standardize data (centered and scaled) using repeated cross validation (5 folds, 3 repeats) to determine the best parameters $\text{n.trees} = {25, 50, ..., 200}$, $\text{interaction.depth} = {1, 2, 3, 4}$, $\text{shrinkage} = 0.1$ and $\text{nminobsinnode} = 10$.

```{r}
# train a boosted tree model with cv and apply the best model on test set
cfm_boost <- fit_model(
  HeartDisease ~ ., df_train, df_test, "gbm", 
  trControl = trainControl(method = "repeatedcv", number = 5, repeats = 3),
  tuneGrid = expand.grid(
    n.trees = seq(25, 200, 25),
    interaction.depth = 1:4,
    shrinkage = 0.1,
    n.minobsinnode =10),
  plot = FALSE, verbose = FALSE)
```

## Comparison

```{r}
tibble(
    kNN = cfm_knn$overall, 
    Tree = cfm_tree$overall,
    Bagged = cfm_bag$overall,
    RandomForest = cfm_rf$overall,
    Boosted = cfm_boost$overall
  ) %>% 
  mutate_all(round, 4) %>% 
  bind_cols(Metric = names(cfm_knn$overall)) %>% 
  relocate(Metric)
```


